---
title: "SMekala_Assign13"
author: "Sekhar Mekala"
date: "Sunday, November 08, 2015"
output: html_document
---

##Problem - 1
Given that $f(x) = x^3 + 2x^2$. 

You can find the derivative of $f(x) = x^3 + 2x^2$ directly also (using the derivative formula $\frac{d(x^n)}{dx} = nx^{n-1}$) as $f'(x) = 3x^2 + 4x$.

We can use the function $f'(x) = 3x^2 + 4x$  to find the derivative of $f(x) = x^3 + 2x^2$ at any given $x$. But we will use the numerical derivative formula to find the derivative, using the following mathematical definition of derivative:

$$f'(x) \approx \frac {f(x + \delta) - f(x)}{\delta}$$
where $\delta$ is very close to zero.

We will build an R function called **appx_derivative()**, using the formula $f'(x) \approx \frac {f(x + \delta) - f(x)}{\delta}$. To verify our work, we will also build another function **exact_derivative()**. The **exact_derivative()** is based on the functional form of the derivative $f'(x) = 3x^2 + 4x$.  Each of these functions will accept a number as input and returns the derivative value of the function $f(x) = x^3 + 2x^2$ at the given input value. Since the function is polynomial, it is continuous for all real numbers, and hence we need not check if the derivative actually exists at a given value.

###R Code for **appx_derivative()** function
We will assume $\delta$ as 1e-6 to compute the derivative of the function at a given value. This approach will give an approximate value of the derivative, at a given value. 

```{r}
appx_derivative <- function(a){
       b = a + 1e-6

       return (((b^3+2*b^2 - (a^3 + 2*a^2))/1e-6))
              
  }
```

###R Code for **exact_derivative()** function
We will use the derivative obtained using the derivative formula ($f'(x) = 3x^2 + 4x$) to implement this function.

```{r}
exact_derivative <- function(a){
       
       return(3*a^2+4*a)
              
  }
```


###Test cases:
We will use the **appx_derivative()** function for certain numbers, and check how close are we to the acutal answer (using **exact_derivative()**) function: 

```{r}
appx_derivative(3)
exact_derivative(3)
appx_derivative(-3)
exact_derivative(-3)
appx_derivative(399.111)
exact_derivative(399.111)
appx_derivative(399)
exact_derivative(399)
```
The above test cases show that our approximation to derivative closely resembles the actual derivative value obtained using the functional form of derivative.



##Problem - 2

Given that $f(x) = 3x^2 + 4x$. We need to find the following:
$$\int_{1}^{3} 3x^2 + 4x dx$$ 

Let us plot this function, in the given interval:
```{r}
curve(3*x^2 + 4*x,from=0,to=5,type="l" )

x <- seq(from=1,to=3,length=10)
y_low <- rep(0,10)
y_high <- 3*(x^2) + 4*x

lines(x,y_low,col='grey')
lines(x,y_high,col='grey')

polygon(c(x, rev(x)), c(y_high, rev(y_low)),
     col = "blue", border = NA)
     

```

We need to determine the shaded area size. Let us assume that this area is divided into many rectangles with the smallest width (say 1e-6), and the length of the rectangle is given by the function $3x^2 + 4x$. The sum of the areas of all such rectangles will give us the area of the shaded region, shown above.

Area of a rectangle is given by LxW (where L = length and W = width)

Therefore the area of the shaded region is given by 
$$\sum (3x^2+4x) . \delta \mbox{ where }x \in [1,3] \mbox{ and } \delta \mbox{ is a very small value, representing the width of the rectangle} $$

The following R code will evaluate this integral by the summation of all the areas of the rectangles having a width of 1e-6 and length of $3x^2+4x \mbox{where } x \in [1, (1+1^{-6}), (1+(2.10^{-6})),(1+(3.10^{-6})), .... 3]$.

```{r}
x = 1
area = 0
delta = 1e-6
while(x<=3)
  {
  
  area <- ((3*x^2+4*x) * delta) + area
  x <- x + delta
  
  }

print(area)
```

Hence the integral evaluates to 41.99998, which is approximately equal to 42.

##Problem - 3
Evaluate $$\int sin(x) cos(x) dx$$

**Method-1**

Let $t = sin(x)$. Differentiating on both sides with respect to $x$...
$$\frac{dt}{dx} = cos(x)$$
$$dt = cos(x) dx$$

Substituting $dt$ in the place of $cos(x) dx$ and $t$ in the place of $sin(x), will make the given integral as 
$$\int sin(x) cos(x) dx = \int (t) dt$$
This evaluates to $$\frac {t^2}{2} + C$$
Substituting back $sin(x)$ in the place of $t$, will give us $$\frac {sin^2(x)}{2} + C$$

Hence, $$\int sin(x) cos(x) dx = \frac {sin^2(x)}{2} + C$$


**Method-2**

Integrating by parts:
$$\int f(x) g(x) dx = f(x) \int g(x) dx - \int f'(x) \int g(x) dx$$

Hence,
$$\int sin(x) cos(x) dx = sin(x) \int cos(x) dx - \int \frac {d(sinx)}{dx} \int cos(x) dx$$
$$sin(x) sin(x) - \int cos(x) sin(x) dx$$

Hence $\int sin(x) cos(x) dx = sin^2(x) - \int cos(x) sin(x) dx + C$ 
This can be re-written as
$$2 \int sin(x) cos(x) dx = sin^2(x) + C$$
Hence, $$\int sin(x) cos(x) dx = \frac {sin^2(x)}{2} + C$$



##Problem - 4
Evaluate $$\int x^2 e^x dx$$
Integrating by parts,
$$\int x^2 e^x dx = x^2 \int e^x dx- \int 2x \int e^x dx = x^2 e^x - 2 \int x e^x dx = x^2 e^x - 2 [x \int e^x dx - \int 1 . \int e^x dx] = x^2 e^x - 2x e^x + 2e^x +C$$

Therefore $$\int x^2 e^x dx = x^2 e^x - 2x e^x + 2e^x +C$$

##Problem - 5
Evaluate $$\frac{d(xcos(x))}{dx}$$

$$\frac{d(xcos(x))}{dx} = \frac{d(x)}{dx} cos(x) + x \frac{d(cox(x))}{dx} = cos(x) + x (-sin(x)) = cos(x) - xsin(x)$$

##Problem - 6
Evaluate $$\frac{d(e^{x^{4}})}{dx}$$
$$\frac{d(e^{x^{4}})}{dx} = e^{x^{4}} . 4. x^3 = 4x^3 e^{x^{4}}$$


                                      ~~~**End of Assignment**~~~