---
title: "SMekala_Assign6"
author: "Sekhar Mekala"
date: "Saturday, September 26, 2015"
output: html_document
---


##Problem set 1

###(1)
When we roll a fair die  3 times, then we will have $6^3$ possible number of outcomes. When we roll the die once we will have any of the 6 possible outcomes {1,2,3,4,5,6}. If we roll the die two times, then we will have {(1,1),(1,2),(1,3),(1,4),(1,5),(1,6),(2,1),(2,2),(2,3),(2,4),(2,5),(2,6),(3,1),(3,2),(3,3),(3,4),(3,5),(3,6),(4,1),(4,2),(4,3),(4,4),(4,5),(4,6),(5,1),(5,2),(5,3),(5,4),(5,5),(5,6),(6,1),(6,2),(6,3),(6,4),(6,5),(6,6)} = $6^2$ = 36 outcomes. Following the same logic, when we roll the die 3 times, then we will have $6^3$ = 216 possible outcomes.

###(2)
When we roll a die 2 times, then we will have $6^2$ = 36 outcomes. To find the probability of getting a sum of 3 (over 2 rolls), we need to identify the favourable number of cases. We will obtain a sum of 3, in 2 cases: {(1,2),(2,1)} = 2 outcomes

Therefore the probability of getting a sum of 3, when a die is rolled 2 times = $\frac{2}{36}$ = $\frac{1}{18}$

###(3)
Give that there are 25 people in a room. The probability that at least 2 of them have the same birthday = 1 - probability that none of them have the same birthday.

Probability that none of them have the same birthday = $\displaystyle \frac{\left( \begin{array}{ccc} 365 \\ 1 \end{array} \right)}{365} . \frac{\left( \begin{array}{ccc} 364 \\ 1 \end{array} \right)}{365} . \frac{\left( \begin{array}{ccc} 363 \\ 1 \end{array} \right)}{365} ...\frac{\left( \begin{array}{ccc} 341 \\ 1 \end{array} \right)}{365} = \prod_{n=1}^{n=25} \frac{\left( \begin{array}{ccc} (365-n + 1) \\ 1 \end{array} \right)}{365} = \frac{P_{25}^{365}}{365^{25}}$   

Therefore, the probability that at least 2 people out of 25 people have the same birthday = $1 - \frac{P_{25}^{365}}{365^{25}}$

The following R code computes this:

```{r}
1 - (choose(365, 25) * factorial(25) / (365 ^ 25))
```

Hence there is approximately 56.87% probability that at least two people out of 25 people have the same birthday.

**NOTE:** The above calculation can be approximated using the following generalized situation:

If we have K balls, and M buckets, and if the balls are thrown at random, then the probability that at least 2 balls go into the same bucket, is approximated by $1-e^{\frac{-k^2}{2M}}$

If we appliy the same approximation to our problem (25 balls translates to 25 people, and 365 buckets translates to 365 days), then we will obtain the approximate probability that at least 2 people have the same birthday as $1-e^{\frac{-25^2}{(2) . (365)}} = 0.5752117$

This value is appoximately equal to the actual value of 56.87%.

If we have 50 people, then the probability of at least 2 people having the same birthday = $1 - \frac{P_{50}^{365}}{365^{50}}$

The following R code finds this:

```{r}
1 - (choose(365, 50) * factorial(50) / (365 ^ 50))

```
Hence there is 97% (appx) chance that at least 2 people (out of 50 people) have the same birthday.

Using our simplified formula, we obtain this probability as $1-e^{\frac{-50^2}{730}} = 0.9674396$

Hence both the probability using the formula ($1-e^{\frac{-k^2}{2M}}$) is approximately equal to the actual probability of 0.9703736









```{r}
library("stringr")

#filetext <- read(file("assign6.sample.txt", encoding = "UCS-2LE"))



#x <- scan(file="assign6.sample.txt", blank.lines.skip = TRUE, multi.line = TRUE, what=character())
#x <- scan(con <- file("assign6.sample.txt", encoding = "UTF-8"), blank.lines.skip = TRUE, multi.line = TRUE, what=character())
x <- scan(file="assign6.sample.txt", encoding = "UTF-8", blank.lines.skip = TRUE, multi.line = TRUE, what=character(),quote="'\"\U2018\U2019")
#x <- scan(file="assign6.sample.txt",  encoding = "cp1252",blank.lines.skip = TRUE, multi.line = TRUE, what=character(),quote="'\"\U2018\U2019")

#Encoding(x) <- "UTF-8"
#close(con)
x <- tolower(x)

clean_data <- function(c)
  {
  
  #Encoding(c) <- "latin1"

  #print(typeof(c))
  #print(class(c))
  
  #print(c)
  #print(c)
  #c <- str_sub(c, start = 1, end = (str_locate(c, "\U2018"))[1][1] - 1)
  #c <- str_sub(c, start = 1, end = (str_locate(c, "\U2019"))[1][1] - 1)
  #print(nchar(c))
  #chk <- str_locate(c, "’", useBytes=T)
  #chk <- str_locate(c, "\u0092")
  #c <- substr(c, start = 1, stop = ifelse(is.na(chk), nchar(c), chk[1][1] - 1))
  #chk <- str_locate(c, '\”')
  #chk <- str_locate(c, '\u0094')
  #c <- substr(c, start = 1, stop = ifelse(is.na(chk), nchar(c), chk[1][1] - 1))

  #chk <- str_locate(c, "'")
  #c <- substr(c, start = 1, stop = ifelse(is.na(chk), nchar(c), chk[1][1] - 1))
  #chk <- str_locate(c, '"')
  #c <- substr(c, start = 1, stop = ifelse(is.na(chk), nchar(c), chk[1][1] - 1))

  #print(c)
  #print(chk)  
  #c <- str_sub(c, start = 1, end = (str_locate(c, "\U2019"))[1][1] - 1)

  #substr(c, start = 1, end = ifelse(is.na(chk), -1, chk[1][1] - 1))
  #c <- str_trim(chartr(".,\"\U201C\U201D)(","       ",c))
  #Encoding(c) <- "UTF-8"
  return(str_trim(str_replace_all(c, "[^[:alnum:]]", "")))
  }

v <- "don’t"
print(clean_data("don’t"))
print(clean_data(" change.” "))
print(clean_data(x[1357])) 
Encoding(x[1357])
typeof("sekhar")
print(clean_data(as.character((x[1357]))))
clean_data(v)
y <- lapply(x, clean_data)
y <- unlist(y)

y <- data.frame(words = y)
as.data.frame(prop.table(table(y)))
as.data.frame(margin.table(table(y),1))
sum(as.data.frame(prop.table(table(y)))$Freq)

y[1][1]
names(y) <- c("z")
print (x)


x[which(x == "for")+1]
```