---
title: "SMekala_Assign6"
author: "Sekhar Mekala"
date: "Saturday, September 26, 2015"
output: html_document
---


##Problem set 1

###(1)
When we roll a fair die  3 times, then we will have $6^3$ possible number of outcomes. 

**Explanation**: When we roll the die once we will have any of the 6 possible outcomes {1,2,3,4,5,6}. If we roll the die two times, then we will have {(1,1),(1,2),(1,3),(1,4),(1,5),(1,6),(2,1),(2,2),(2,3),(2,4),(2,5),(2,6),(3,1),(3,2),(3,3),(3,4),(3,5),(3,6),(4,1),(4,2),(4,3),(4,4),(4,5),(4,6),(5,1),(5,2),(5,3),(5,4),(5,5),(5,6),(6,1),(6,2),(6,3),(6,4),(6,5),(6,6)} = $6^2$ = 36 outcomes. Following the same logic, when we roll the die 3 times, then we will have $6^3$ = 216 possible outcomes.

###(2)
When we roll a die 2 times, then we will have $6^2$ = 36 outcomes. To find the probability of getting a sum of 3 (over 2 rolls), we need to identify the favourable number of cases. We will obtain a sum of 3, in 2 cases: {(1,2),(2,1)} = 2 outcomes

Therefore the probability of getting a sum of 3, when a die is rolled 2 times = $\frac{2}{36}$ = $\frac{1}{18}$

###(3)
Given that there are 25 people in a room. **The probability that at least 2 of them have the same birthday = 1 - probability that none of them have the same birthday**.

Probability that none of them have the same birthday = $\displaystyle \frac{\left( \begin{array}{ccc} 365 \\ 1 \end{array} \right)}{365} . \frac{\left( \begin{array}{ccc} 364 \\ 1 \end{array} \right)}{365} . \frac{\left( \begin{array}{ccc} 363 \\ 1 \end{array} \right)}{365} ...\frac{\left( \begin{array}{ccc} 341 \\ 1 \end{array} \right)}{365} = \prod_{n=1}^{n=25} \frac{\left( \begin{array}{ccc} (365-n + 1) \\ 1 \end{array} \right)}{365} = \frac{P_{25}^{365}}{365^{25}}$   

Therefore, the probability that at least 2 people out of 25 people have the same birthday = $1 - \frac{P_{25}^{365}}{365^{25}}$

The following R code computes this:

```{r}
1 - (choose(365, 25) * factorial(25) / (365 ^ 25))
```

Hence there is approximately **56.87%** probability that at least two people out of 25 people have the same birthday.

**NOTE:** The above calculation can be approximated using the following generalized situation:

If we have **K balls**, and **M buckets**, and if the balls are thrown at random into buckets, then the probability that at least 2 balls go into the same bucket, is approximated by $1-e^{\frac{-k^2}{2M}}$

If we appliy the same approximation to our problem (25 people translates to 25 balls, and 365 days translates to 365 buckets), then we will obtain the approximate probability that at least 2 people have the same birthday as $1-e^{\frac{-25^2}{(2) . (365)}} = 0.5752117$

This value is **appoximately** equal to the actual value of **56.87%**.

If we have 50 people, then the probability of at least 2 people having the same birthday = $1 - \frac{P_{50}^{365}}{365^{50}}$

The following R code finds this:

```{r}
1 - (choose(365, 50) * factorial(50) / (365 ^ 50))

```

Hence there is **97% (appx)** chance that at least 2 people (out of 50 people) have the same birthday.

Using our simplified formula, we obtain this probability as $1-e^{\frac{-50^2}{730}} = 0.9674396$

Hence both the probability using the formula ($1-e^{\frac{-k^2}{2M}}$) is approximately equal to the actual probability of **0.9703736**



##Problem set 2

**NOTE**:

* Please install "stringr" package, as the following R code needs that package to successfully run.

* Please copy that file (**assign.sample.txt**) to the current working directory, from where you are building this RMD document.

The R code to read the file **assign6.sample.txt** follows. 

The following R code will read the file, and converts all the text to lower case.

```{r}
library("stringr")

#Reading the file into "data" variable. The file's data is in UNICODE, so using UTF-8 format to read the file
data <- scan(file="assign6.sample.txt", encoding = "UTF-8", blank.lines.skip = TRUE, multi.line = TRUE, what=character())

#Converting the read data to lower case
data <- tolower(data)

```

The following R code will create a function named **clean_data**, which removes all the non-alphanumeric letters from the words.

```{r}
#The clean_data function removes all quotes, hyphens all all non-alphanumeric characters

clean_data <- function(c)
{
  
  return(str_trim(str_replace_all(c, "[^[:alnum:]]", "")))
}

```


The following R code will calculate and displays the probability of each word occurring in the document (assign6.sample.txt)

```{r}

unigram <- lapply(data, clean_data)

unigram <- unlist(unigram)

unigram <- unigram[unigram != ""]

unigram_df <- data.frame(words = unigram)

unigram_prob <- as.data.frame(sort(prop.table(table(unigram_df)),decreasing=T))
names(unigram_prob) <- c("Probability")
print(unigram_prob)

```

Now let us find the probabilities of 2 words occurring together. The following R code gets the probabilities of all the bigrams in the given text (only the top 40 bigrams order by descending probability are displayed):

```{r}
#Finding bigram probabilities

unigram_temp <- c(unigram[-1], ".")

#Finding all bigrams using paste

bigram <- paste(unigram, unigram_temp)

#sum((bigram == "eyes the"))/sum(unigram == "the")

bigram_df <- data.frame(words = bigram)

bigram_prob <- as.data.frame(sort(prop.table(table(bigram_df))),decreasing = T)
names(bigram_prob) <- c("Probability")
head(bigram_prob,40)

```

##Test cases

###Test case 1
Let us find the probability of two words **for, the** occuring together. From the bigrams probability (present in **bigram** vector), we can find the probability of having the word **the** after **for**. Next we will get the probability of having **for** after **the**. Then the sum of these probabilities will give probability of having these two words together in the document. 

Mathematically, we can say

$$P((for, the) \cup (the, for)) = P(for, the) + P(the, for) - P((for, the) \cap (the, for))$$

But $P((for, the) \cap (the, for))$ will be zero, since a two word combination cannot have $(for, the)$ and $(the,for)$ at the same time.

Therefore $P((for, the) \cup (the, for)) = P(for, the) + P(the, for)$

Probability of "the" immediately after "for" (represented as $P(for, the)$) is obtained using the following R code:

```{r}
sum((bigram == "for the"))/sum(unigram == "for")

```


On the similar lines, the probability of having "for" immediately after "the" (represented as $P(the, for)$) is obtained using the following R code:

```{r}
sum((bigram == "the for"))/sum(unigram == "the")
```

Finally, the probability of "for" and "the" occurring together will be $P((for, the) \cup (the, for)) = P(for, the) + P(the, for) = 0.1290323 + 0 = 0.1290323$

The probabilities of occurring "the" and "for" (as single words) is found below:

```{r}
#Probability of "the"
sum(unigram == "the")/length(unigram)

#Probability of "for"
sum(unigram == "for")/length(unigram)

```

Let us check if "for" and "the" occurrence is independent:

```{r}
#Probability of "the"
(sum(unigram == "the")/length(unigram)) * (sum(unigram == "for")/length(unigram))

```
The above probability is NOT the same as the probability we found for $P("for" \cap "the")$. Hence these two words are NOT independent.


###Some more test cases

Following the same logic, here are the probabilities of some words occurring together in the document: "said" and "were", "can" and "change", "prisons" and "problems", "that" and "is", "favors" and "she"


```{r}

sum((bigram == "said were"))/sum(unigram == "said") + sum((bigram == "were said"))/sum(unigram == "were")

sum((bigram == "can change"))/sum(unigram == "can") + sum((bigram == "change can"))/sum(unigram == "change")

sum((bigram == "prisons problems"))/sum(unigram == "prisons") + sum((bigram == "problems prisons"))/sum(unigram == "problems")

sum((bigram == "that is"))/sum(unigram == "that") + sum((bigram == "is that"))/sum(unigram == "is")

sum((bigram == "favors she"))/sum(unigram == "favors") + sum((bigram == ""))/sum(unigram == "is")

```

                                              ~~~** End of Home Work **~~~